Landcover Data Limits Our Understanding of Global Change

### 2015-11-25 paper/main.tex
Near-final version of main text, pending Tom's edits. Confusion matrix added to Appendix.

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,sort,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}
%\usepackage{placeins}
%\usepackage{float}
%\restylefloat*{figure}
%\usepackage{dblfloatfix}  % http://tex.stackexchange.com/questions/235623/placing-a-figure-in-the-bottom-of-a-page-spanning-the-two-columns-of-an-ieee-doc 
%\usepackage{subfigure}
%\usepackage{wrapfig}
%\usepackage{caption}
%\setlength\belowcaptionskip{-2cm}
%\setlength{\textfloatsep}{5pt plus 1.0pt minus 2.0pt}
\setlength{\parskip}{1cm plus4mm minus3mm}

@@

\begin{document}

@@

\author{Lyndon Estes\affil{1}{Princeton University, Princeton, NJ USA},
Peng Chen\affil{2}{Indiana University, Bloomington, IN USA},
Stephanie Debats\affil{1}{Princeton University, Princeton, NJ USA},
Tom Evans\affil{2}{Indiana University, Bloomington, IN USA},
Fanie Ferreira\affil{3}{GeoTerraImage, Pretoria, RSA},
Gabrielle Ragazzo\affil{1}{Princeton University, Princeton, NJ USA},
Justin Sheffield\affil{1}{Princeton University, Princeton, NJ USA}
Adam Wolf\affil{1}{Princeton University, Princeton, NJ USA}
\and
Kelly Caylor\affil{1}{Princeton University, Princeton, NJ USA}}

@@

\begin{article}
\begin{abstract}
{Our knowledge of global change processes is built upon landcover data, but we lack the means to comprehensively assess landcover maps, particularly in rapidly developing regions, and thus do not fully comprehend how map quality impacts our understanding of global change. We used a unique, high quality, national-scale cropland map to assess the bias and accuracy in a) current generation landcover products, and b) analyses built on these. Maps have high bias and low accuracy, particularly at finer resolutions (1-10 km), which transfers into, and is often amplified, in downstream analyses of vegetative carbon stock, gridded crop production, and process-based food security models. Substantial map aggregation is needed to avoid potential misunderstandings of global change processes or misguided policy. Maps derived from coarse resolution sensors (such as MODIS and GlobCover) should be aggregated to at least 50-100 km resolution (and sometimes more). Map accuracy is lowest at intermediate levels of cover, thus coarser aggregation may be necessary in highly mixed landscapes. Higher resolution imaging systems (e.g. Landsat) produce maps that require less aggregation (1-25 km) to achieve high accuracy and low bias. For regional to global analyses, new fusion products, such as the GeoWiki cropland map, need just 5 km of aggregation to produce unbiased maps for most applications, but can require $\geq$25 km of aggregation to have sufficient accuracy. Widely used global datasets for carbon, crop area, and crop yield should be rebuilt on GeoWiki and similar maps, while new approaches methods that fuse human judgement, the latest computer vision-based algorithms, and growing high resolution image archives are needed to produce the next generation of landcover maps. }
\end{abstract}

@@

\abbreviations{GTI, GeoTerraImage; SSA, sub-Saharan Africa}
\linenumbers

@@

The value of the insights resulting from such studies depends upon the veracity of their underlying landcover data, much as a house requires a solid foundation in order to remain standing. Unfortunately, the evidence to date indicates that much of our understanding of global change is built on shaky foundations. The reason for this is that landcover data can only practically be derived from satellite imagery, and in many regions the cover types of interest are smaller \cite[e.g. smallholder's farms][]{jain_mapping_2013} than the sensor resolution, or spectrally indistinct from neighboring covers, which translates into substantial mapping errors \cite{see_improved_2015,lobell_use_2013,estes_diylandcover:_2015}. Landcover maps are therefore often inaccurate at finer scales and disagree widely between products, particularly in the world's most rapidly developing regions \cite{estes_projected_2013, fritz_comparison_2010, fritz_cropland_2011}. These errors mean that we are still unable to obtain the granular, mechanistic understanding of global change processes that we need.

These problems with landcover products are known \cite{fritz_comparison_2010, fritz_cropland_2011, see_improved_2015, fritz_mapping_2015,verburg_challenges_2011}, and there are a variety of map improvement efforts underway, particularly for agriculture \cite{fritz_geo-wiki:_2012,estes_diylandcover:_2015}. What remains an open question is exactly how much the maps researchers typically use deviate from actual landcover, and how this in turn impacts our understanding of global change processes. Answering this question depends on having spatially comprehensive ground truth data, which are unavailable for most parts of the world, particularly over Africa and other developing regions \cite{see_improved_2015}. Our understanding of map accuracy is therefore built primarily on bottom-up tests made with a relatively small number of ground truth points (relative to the total mapped area), or from top-down "sanity checks" made in comparison to aggregated survey data. This allows us to quantify between-map discrepancies  \cite[e.g.][]{fritz_comparison_2010, kaptue_tchuente_comparison_2011}, or to understand map fidelity to country-level statistics \cite[e.g.][]{fritz_comparison_2010}, but offers little direction for how to arrive at a true number.

Being unable to fully quantify the errors in landcover maps makes it difficult, if not impossible, to gauge their impact on downstream analyses. There has been some work examining how such error influences climate simulations \cite{ge_impacts_2007}, agricultural land use patterns \cite{schmit_limitations_2006}, carbon flux measurements \cite{quaife_impact_2008}, and human population estimates \cite{linard_assessing_2010}, but these either use simulated landcover errors \cite{ge_impacts_2007} or compare relevant differences in estimates between different satellite-derived landcover maps \cite{linard_assessing_2010, quaife_impact_2008}. One exception is a Belgian study \cite{schmit_limitations_2006} that used ground-collected farm parcel data to assess how landcover errors bias measurements of agricultural land use patterns, but the study extent was fairly small and the validation data were discontiguous.

Fortunately, the recent, explosive growth in public and private initiatives to develop new Earth observing capabilities, which range from small drones\footnote{e.g. 3DRobotics, DJIA} to new high resolution satellite arrays\footnote{Planet Labs, Skybox} and better mapping methods \cite{fritz_geo-wiki:_2012,estes_projected_2013,debats_generalized_????}, are finally providing the means to comprehensively interrogate the accuracy and biases in the landcover products that have become commonplace in global change research--and which are often used to make policy decisions \cite{searchinger_high_2015}.

In this study, we take advantage of these recent advances to address the urgent need to more thoroughly and precisely quantify landcover map errors and how they might impact our understanding of change in the world's most dynamic regions.  We use a unique, high accuracy landcover map of South African crop fields to conduct a spatially comprehensive, bottom-up quantification of error in several widely used landcover maps, and how these errors can propagate through ``downstream'' studies investigating into both the physical and socioeconomic components of global change. Our objective is to provide scientists and policy-makers who use landcover data with a better, more up-to-date, understanding of their appropriate uses and limitations.

\vspace{-0.5 cm}
\section{Overview of study area and analyses}
In the late 2000s, the South African government commissioned a cropland map that was made by manually interpreting and digitizing fields visible within high resolution satellite imagery \cite{fourie_better_2009}. The resulting vectorized field boundaries provide highly accurate data on field sizes and distribution for the period 2009-2011. This dataset is particularly valuable because South Africa represents nearly 6\% of sub-Saharan Africa's (SSA) area, which is a region that is poised to undergo rapid agricultural expansion \cite{searchinger_high_2015}, yet is notably lacking trustworthy maps of existing agriculture land \cite{fritz_comparison_2010}. Moreover, South Africa's large agricultural sector represents the diversity of farming systems found throughout SSA, ranging from large commercial operations to smallholder farms \cite{hardy_rainfed_2011,estes_using_2014}.

We used this dataset as a reference layer for evaluating four landcover products representative of the type commonly used in Earth Systems research. The first was South Africa's own 30 m resolution 2009 National Landcover map (SA-LC)\cite{sanbi_national_2009}, which is typical of the higher-resolution, Landsat-based maps that are typically available only for individual countries \cite[e.g.][]{fry_completion_2009}. The second and third were respectively the 300 m GlobCover 2009 \cite{arino_global_2012} and 500 m resolution 2011 MODIS Landcover products, which are widely used global-scale products \cite[e.g.][]{gross_monitoring_2013, shackelford_conservation_2015}. The fourth dataset was the new 1 km GeoWiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}, which incorporates the first three datasets and represents the current state-of-the-art in landcover mapping.  For comparison, we converted all five datasets into 1 km gridded maps of cropland density, expressed as the pixel-wise percent cover.

In our first set of comparisons, we evaluated map quality by subtracting each of the four landcover product-derived percent cropland maps (hereafter we refer to these as the ``the test maps'') from the reference map, so that we could calculate each map's bias (the mean pixel-wise error) and accuracy (the mean absolute pixel-wise error, where a lower value indicates higher accuracy). We performed this analysis for the original 1 km resolution, and for maps that were further aggregated to 5, 10, 25, 50, and 100 km, in order to evaluate how error and bias changes with scale. We also assessed how landcover pattern impacts map performance by modeling the correlation between map accuracy and cropland density.

@@

Our reference map indicated that crop fields covered 104,304 km$^2$, or nearly 10\%, of the total mapped area in the 2009-2011 time period. The test maps derived from SA-LC and GeoWiki overestimated this area by 31 and 10\%, respectively, while GlobCover and MODIS underestimated it by 18 and 23\%. Subtracting each test map from the reference maps created pixel-wise residuals, where negative and positive values respectively represent overestimates and underestimates by the test map (Fig. 1A).

\begin{figure*}[t]
%\begin{wrapfigure}{c}{1\textwidth}
\centerline{\includegraphics[width=0.95\textwidth]{figures/figure1.pdf}}
\vspace{-0.15 cm}
\caption{(A) Errors in the percent cropland estimates resulting from each of the four test maps relative to the reference map at different scale of pixel aggregation. Rows indicate the test map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas where areas under communal farmlands or permanent tree crops were removed from analysis. (B) The bias (mean error) and accuracy (mean absolute error [MAE]) of each test map at each scale of aggregation, weighted by the percentage cropland in each cell of the reference map. Bias estimates are indicated by the semi-transparent bars, accuracy (lower is more accurate) by the solid bars, with bar colors coded to specific cropland maps.}
\label{afoto1}
\end{figure*}
%\end{wrapfigure}

%\begin{figure*}[htp]
%\centering
%    \subfigure[random caption 1]{\includegraphics[scale=0.49]{figures/bias_map.pdf}}\quad
%     \subfigure[random caption 1]{\includegraphics[scale=0.49]{figures/bias_map.pdf}}
%\end{figure*}

\linenumbers
The most pronounced errors were in the MODIS and GlobCover maps, which showed large positive residuals in the center of the country where cropland is most concentrated (blue areas in Fig. 1A), and negative residuals (red areas) along the eastern and northern margins.
These patterns translated into substantial map bias (Fig. 1.B), with GlobCover and MODIS mean error (weighted by the reference cropland density) exceeding 45\% and 25\% respectively at 1 km resolution, meaning that each map tends to underestimate cropland by that amount at that resolution. This bias declined with each level of map aggregation, being reduced to nearly 15\% for GlobCover and 5\% for MODIS at 100 km. The magnitude of (density weighted) mean absolute error (MAE) was somewhat higher in all cases. The GeoWiki map, in contrast, was the least biased overall, showing just a ~7\% bias at 1 km and near 0 for all other scales of aggregation, although its accuracy (23\% MAE) was only half as good as SA-LC's at 1 km (11\% MAE), which despite its uniform overestimation bias (Fig. 1A) was the most accurate map at aggregation scales $<10$km. Above this, GeoWiki became slightly more accurate, having $<$5\% MAE at 100 km resolution. The reason GeoWiki had relatively poor accuracy at 1 km resolution was due to the heterogeneity of residuals, which traded between positive and negative residuals over short distances, thereby inflating MAE at this scale.

\subsection{Error and landcover density}
Given that landcover characteristics can influence landcover map quality \cite{see_improved_2015, estes_diylandcover:_2015, debats_generalized_????}, we set out to elucidate the relationship between cropland configuration and map error.  We first calculated the average 1 km reference cropland percentage within each of South Africa's 354 magisterial districts (the finest administrative unit, averaging 3,445 km$^2$; SI), which yielded a landscape-scaled metric of characteristic cropland density, and similarly calculated the district-wise MAE for each test map.

@@

The magnitude of carbon errors varied as a function of the carbon density of surrounding cover, as demonstrated by the bias statistics (Fig. 3). Bias was near zero when grassland was the adjacent cover type (SI), as its carbon density is nearly the same as cropland. However, when forest was adjacent then bias was a three- to five-fold multiple of cropland map bias (Fig. 1B). At the most extreme, GlobCover's bias was -276\% at 1 km, but even SA-LC and GeoWiki had biases of 22\% and -50\%, respectively. Bias could be substantial even for the least carbon dense vegetation type (sparse), as evidenced by the 15-25\% bias at 1 km for MODIS and GlobCover under this class.  The mean bias across the different potential adjacent vegetation classes ranged between -20 for GeoWiki and -123\% for GlobCover at 1 km (with MODIS in between these), while SA-LC's average bias was 11\%.  Biases declined fairly rapidly with aggregation, with all datasets having an average (across cover types) bias magnitude of less than 10\% by 25 km of aggregation, except for GlobCover, which was -12\% at 100 km (SI).  As with cropland percentages, GeoWiki produced the least biased carbon density estimates above above 1 km resolution.

In terms of accuracy, MAE values were essentially the same as bias magnitudes, except for GeoWiki's, which were twice as large. The average MAE across vegetation classes was 47\% at 1 km, dropping to $<$10 only with 25 km of aggregation. In contrast, SA-LC's carbon estimates were twice as accurate at 1 km, and was slightly more accurate up to 25 km of aggregation were GeoWiki achieved parity.

@@

We used these same methods  \cite{ramankutty_farming_2008,monfreda_farming_2008} to disaggregate maize harvested area (South Africa's largest crop \cite{estes_projected_2013}) on top of our reference and cropland maps, followed by yields, which were assigned to cells have harvested areas greater than zero. We then used these two layers to calculate maize production, and further aggregated the yield and production grids to 5, 10, 25, 50, and 100 km resolutions before quantifying the bias and accuracy of each test map's yield and production values. In this case, we could not convert cell-wise errors into percentages of the reference map values (because many cells had zero values for one map but not the other), so calculated bias and accuracy from the map residuals and then normalized their values to the reference map means.

Maize yields disaggregated onto the test maps showed some marked differences relative to the reference map, but only at the margins of the major crop production areas where cropland is sparser (SI). These differences resulted when a yield value was mapped onto a grid cell where the reference map had no harvested area, and thus zero yield. In more densely cropped areas, such discrepancies were less frequent because both the reference and test maps were both likely to have some maize harvested area, and therefore a yield value.  Yield biases were thus fairly low (and accuracy high), with the largest being 20\% for MODIS at 1 km, following by GlobCover with 10\% (Fig. 5). These dropped to below 10\% with aggregation.

Production biases were generally slightly higher, but still low, for most datasets, with the exception of GlobCover, which had an gigantic underestimation bias of $>$60\% (relative to mean production) at 1 km, which remained above 10\% even at 100 km of aggregation. MODIS production bias was above 20\% at 1 km, but declined to below 10\% at higher levels of aggregation.

>>>>

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,sort,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}
%\usepackage{placeins}
%\usepackage{float}
%\restylefloat*{figure}
%\usepackage{dblfloatfix}  % http://tex.stackexchange.com/questions/235623/placing-a-figure-in-the-bottom-of-a-page-spanning-the-two-columns-of-an-ieee-doc 
%\usepackage{subfigure}
%\usepackage{wrapfig}
%\usepackage{caption}
%\setlength\belowcaptionskip{-2cm}
%\setlength{\textfloatsep}{5pt plus 1.0pt minus 2.0pt}
\usepackage{multirow}
\setlength{\parskip}{1cm plus4mm minus3mm}
\usepackage{makecell}
\usepackage{boldline}
\setcellgapes{3pt}
\usepackage[flushleft]{threeparttable}

@@

\begin{document}
%\hspace*{\fill}

@@

\author{Lyndon Estes{\affil{1}{Woodrow Wilson School, Princeton University, Princeton, NJ USA}\affil{2}{Civil and Environmental Engineering, Princeton University, Princeton, NJ USA}},
Peng Chen\affil{3}{Indiana University, Bloomington, IN USA},
Stephanie Debats\affil{2}{},
Tom Evans\affil{3}{},
Stefanus Ferreira\affil{4}{GeoTerraImage, Pretoria, RSA}{},
Gabrielle Ragazzo\affil{5}{Ecology and Evolutionary Biology, Princeton University, Princeton, NJ USA},
Justin Sheffield\affil{2}{},
Adam Wolf\affil{5}{},
Eric Wood\affil{2}{},
\and
Kelly Caylor\affil{2}{}}

@@

\begin{article}
\begin{abstract}
{Research into global environmental change is often built upon landcover maps, which can have substantial errors. How these errors impact our understanding of global change processes and associated policies is thus of great interest, but is difficult to fully quantify because we lack spatially comprehensive reference data, particularly in the World's most rapidly developing regions. We used a high quality, national-scale cropland map to assess bias and accuracy in current generation landcover maps and four examples of ``downstream'' (landcover-dependent) analyses, two related to physical processes (estimates of vegetative carbon stocks and evapotranspiration) and two to socio-economic processes (gridded crop production estimates and agent-based simulation of household food security), at resolutions ranging from 1-100 km. We found that the landcover maps have substantial errors below 25 km resolution, with substantial biases and inaccuracies, particularly from maps derived from coarse resolution sensors. In some cases these error metrics remain high even when maps are aggregated to 100 km. These errors can be substantially amplified in downstream analyses (e.g. the carbon and crop production analyses), but in cases where the values of different cover types that underpin the analysis are similar (e.g. evapotranspiration estimates) the results can be relatively insensitive. To avoid these errors, and thus minimize the risk of misunderstanding global change processes or misinforming policy, substantial map aggregation of landcover maps from their base resolution is often needed. We provide several guidelines that can help landcover map users select appropriate datasets and  aggregation scales under different use cases.

%To achieve low \emph{bias} in a map-wide statistic (e.g. average carbon density), then finer resolutions can be maintained, but if high \emph{accuracy} is desired (e.g. spatial comparisons of map values, or change detection), then coarser pixels are needed.
%Map accuracy is lowest at intermediate levels of cover, thus if the mapped region has highly heterogenous cover (e.g. regions dominated by smallholder-farming), then greater aggregation is also necessary. From a dataset perspective, maps derived from coarse resolution sensors (e.g. MODIS) often need to be aggregated to at \textcolor{red}{50-100} km resolution for bias to be reduced to within +\-10\%, and \textcolor{red}{50-100+} km for mean absolute error (MAE), a measure of accuracy, to be <10\%. For comparable rates of bias and error, higher resolution imaging systems (e.g. Landsat) produce maps that require less aggregation (1-25 km).  For regional to global analyses, new fusion products, such as the GeoWiki cropland map, which constrain maps so that the sum of their pixels match a total reported for a larger scale (e.g. government cropped area statistics for administrative districts), need just 5 km of aggregation to produce unbiased maps for most applications. However, this technique does less to improve accuracy, which may still require $\geq$25 km or more aggregation to achieve <10\% MAE. Widely used global datasets for carbon, crop area, and crop yield should be rebuilt on GeoWiki and similar maps, while new methods that fuse human judgement, the latest computer vision-based algorithms, and growing high resolution image archives can help produce the next generation of landcover maps. 
}
\end{abstract}

@@

%\abbreviations{GTI, GeoTerraImage; SSA, sub-Saharan Africa}
\linenumbers

@@

The value of the insights resulting from such studies depends upon the veracity of their underlying landcover data, much as a house requires a solid foundation in order to remain standing. Unfortunately, the evidence to date indicates that much of our understanding of global change is built on shaky foundations. The reason for this is that landcover data can only practically be derived from satellite imaging, which has several important constraints--related to both physics and analytics--that propagate mapping errors. First, in many regions the cover types of interest are smaller than the sensor resolution \cite[e.g. smallholder's farms][]{jain_mapping_2013,debats_generalized_2015,ozdogan_resolution_2006}, or spectrally indistinct from neighboring covers \cite{sweeney_mapping_2015}. Second, the act of defining a cover class can cause error, in that selected classes may have highly diverse spectral properties (e.g. agriculture \cite{debats_generalized_2015,estes_platform_2015}) and thus be difficult for the classifier to distinguish, while discretizing a continuous cover type (e.g. dividing a forest into different canopy cover classes) can also promote classification error, particularly near class boundaries \cite{foody_status_2002}, as well as confusion about the actual extent of the cover type \cite{sexton_conservation_2015}. Furthermore, class definitions often vary between maps, complicating inter-comparison \cite{kuemmerle_challenges_2013}. Third, landcover map series are often used to detect changes \cite[e.g.][]{gross_monitoring_2013}, or may use within-class seasonality as a means for discriminating between classes (e.g. phenology of cropping versus surrounding vegetation \cite{sweeney_mapping_2015}), but seasonal variability and landcover changes can be easily confused. Given these multiple sources of error, landcover maps are often inaccurate at finer scales and disagree widely between products, particularly in the world's most rapidly developing regions \cite{estes_projected_2013, fritz_comparison_2010, fritz_cropland_2011}. These errors limit our ability to obtain granular, mechanistic understanding of global change processes.

These problems with landcover products are well known \cite{fritz_comparison_2010, fritz_cropland_2011, see_improved_2015, fritz_mapping_2015,verburg_challenges_2011}, and there are a variety of map improvement efforts underway \cite[e.g.][]{fritz_geo-wiki:_2012, estes_platform_2015}. What remains an open question is exactly how much the maps researchers typically use deviate from actual landcover, and how this in turn impacts our understanding of global change processes. Fully answering this question depends on having comprehensive, spatially representative ground truth data, which are unavailable for most parts of the world, particularly over Africa and other developing regions \cite{see_improved_2015}. Our current understanding of map accuracy is therefore built primarily on bottom-up tests made with a relatively small number of ground truth points (relative to the total mapped area), or from top-down "sanity checks" made in comparison to aggregated survey data.  In terms of the former, there are well-developed, best-practice guidelines for selecting and using ground-truth samples to robustly quantify map error, but these are not commonly applied by the broader community of landcover map users (often due to lack of data), which makes it challenging to compare how error varies within and between maps \cite{stehman_global_2012, olofsson_making_2013,olofsson_good_2014,foody_status_2002}. The latter form of assessment are made at coarser administrative levels relative to reported statistics \cite[e.g.][]{fritz_comparison_2010}, which can also be suspect \cite{carletto_emperor_2013,fao_action_2013}.

%Top-down accuracy assessments are also made from  "sanity checks" made in comparison to aggregated survey data. 
%
%
%
%This allows us to quantify between-map discrepancies  \cite[e.g.][]{fritz_comparison_2010, kaptue_tchuente_comparison_2011}, or to understand map fidelity to country-level statistics \cite[e.g.][]{fritz_comparison_2010}, but offers little direction for how to arrive at a true number.
%
%
%
%In practice, this means that we have an incomplete understanding of map error, and 
%
%
%
%
%There are also well-defined guidelines for quantifying and accounting for map accuracy \cite{stehman_global_2012, olofsson_good_2014, olofsson_good_2014}. However, inadequate assessment of map error and their impacts on subsequent analyses remains quite common within the community of landcover map users \cite{olofsson_making_2013,olofsson_good_2014}. There is also an open question  
%
%or from top-down "sanity checks" made in comparison to aggregated survey data.

%This allows us to quantify between-map discrepancies  \cite[e.g.][]{fritz_comparison_2010, kaptue_tchuente_comparison_2011}, or to understand map fidelity to country-level statistics \cite[e.g.][]{fritz_comparison_2010}, but offers little direction for how to arrive at a true number.

Since it is difficult to fully quantify landcover map errors, it is even more challenging to gauge their impact on downstream analyses, in which there is substantial risk of error amplification\cite{kuemmerle_challenges_2013}. There has been some work examining how map error influences climate simulations \cite{ge_impacts_2007}, agricultural land use patterns \cite{schmit_limitations_2006}, carbon flux measurements \cite{quaife_impact_2008}, human population estimates \cite{linard_assessing_2010}, and species distribution models \cite{tuanmu_global_2014}, but these either use simulated landcover errors \cite{ge_impacts_2007} or compare relevant differences in estimates between different satellite-derived landcover maps \cite{linard_assessing_2010, quaife_impact_2008,tuanmu_global_2014}. One exception is a Belgian study \cite{schmit_limitations_2006} that used ground-collected farm parcel data to assess how landcover errors bias measurements of agricultural land use patterns, but the study extent was fairly small and the validation data were discontiguous.

Fortunately, the recent, explosive growth in public and private initiatives to develop new Earth observing capabilities, which range from small drones\footnote{e.g. 3DRobotics, DJIA} to new high resolution satellite arrays\footnote{Planet Labs, Skybox} and better mapping methods \cite{fritz_geo-wiki:_2012,estes_projected_2013, debats_generalized_2015}, are finally providing the means to more comprehensively interrogate the accuracy and biases in the landcover products that have become commonplace in global change research--and which are often used to make policy decisions \cite{searchinger_high_2015}.

In this study, we take advantage of this recent growth in high resolution data to address the call to more thoroughly assess errors and in landcover maps \cite{kuemmerle_challenges_2013, olofsson_good_2014,olofsson_global_2012}, and further examine how these errors might impact our understanding of change in the world's most dynamic regions.  We use a unique, high accuracy landcover map of South African crop fields to conduct a spatially comprehensive, bottom-up quantification of error in several widely used landcover maps, and how these errors can propagate through downstream studies investigating both the physical and socioeconomic components of global change. Our objective is to provide scientists and policy-makers who use landcover maps an up-to-date assessment of their appropriate uses and limitations.

\vspace{-0.5 cm}
\section{Overview of study area and analyses}
In the late 2000s, the South African government commissioned a cropland map that was made by manually interpreting and digitizing fields visible within high resolution satellite imagery \cite{fourie_better_2009}. The resulting vectorized field boundaries provide highly accurate data on field sizes and distribution for the period 2009-2011. Measured in terms of its ability to distinguish crop fields from other cover types, this dataset is 97\% percent accurate at 1 km resolution, with user's accuracies of 94\% and 98\% for the cropland and non-cropland classes, and producer's accuracies of 84 and 99\%  \cite[][SI]{estes_platform_2015}.

This dataset is particularly valuable because South Africa represents nearly 6\% of sub-Saharan Africa's (SSA) area, which is a region that is poised to undergo rapid agricultural expansion \cite{searchinger_high_2015}, yet is notably lacking trustworthy maps of existing agriculture land \cite{fritz_comparison_2010}. Moreover, South Africa's large agricultural sector represents the diversity of farming systems found throughout SSA, ranging from large commercial operations to smallholder farms \cite{hardy_rainfed_2011,estes_using_2014}.

We used this dataset as a reference layer for evaluating four landcover products representative of the type commonly used in Earth Systems research. The first was South Africa's own 30 m resolution 2009 National Landcover map (SA-LC)\cite{sanbi_national_2009}, which is typical of the higher-resolution, Landsat-based maps that are typically available only for individual countries \cite[e.g.][]{fry_completion_2009}. The second and third were respectively the 300 m GlobCover 2009 \cite{arino_global_2012} and 500 m resolution 2011 MODIS Landcover products, which are widely used global-scale products \cite[e.g.][]{gross_monitoring_2013, shackelford_conservation_2015}. The fourth dataset was the new 1 km GeoWiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}, which incorporates the first three datasets and represents the current state-of-the-art in landcover mapping.  Since GeoWiki provides a continuous bounded value of landcover (percent cropland cover), which cannot be feasibly converted to a categorical value, we converted all other datasets, including the reference vectors, into comparable 1 km gridded percent cropland estimates.

We first evaluated the quality of the landcover product-derived percent cropland maps (hereafter referred to as the ``test maps''). Instead of the standard confusion matrix-based accuracy assessment metrics \cite{olofsson_good_2014,olofsson_making_2013}, which apply to categorical landcover maps, we assessed the bias and accuracy of the test maps based on the gridded residuals that resulted when each test map was subtracted from the reference map. Here bias is the mean residual value, weighted by the density of reference cropland, and accuracy is the mean of the absolute values of residuals (also weighted by cropland density), thus accuracy is inversely related to this value. We calculated these metrics for the original 1 km resolution, and for maps that were further aggregated to 5, 10, 25, 50, and 100 km, in order to evaluate how bias and accuracy changes with scale. We also assessed how landcover pattern impacts map performance by modeling the correlation between map accuracy and cropland density.

@@

Our reference map indicated that crop fields covered nearly 10\% of the total study area in the 2009-2011 time period. Subtracting each test map from the reference maps created pixel-wise residuals, where negative and positive values respectively represent overestimates and underestimates by the test map (Fig. 1A).

%\end{wrapfigure}

%\begin{figure*}[htp]
%\centering
%    \subfigure[random caption 1]{\includegraphics[scale=0.49]{figures/bias_map.pdf}}\quad
%     \subfigure[random caption 1]{\includegraphics[scale=0.49]{figures/bias_map.pdf}}
%\end{figure*}
%\linenumbers
The most pronounced errors were in the MODIS and GlobCover maps, which showed large positive residuals in the center of the country where cropland is most concentrated (blue areas in Fig. 1A), and negative residuals (red areas) along the eastern and northern margins.
These patterns translated into substantial map bias (Fig. 1B), with GlobCover and MODIS mean error exceeding 45\% and 25\% respectively at 1 km resolution, meaning that each map tends to underestimate cropland by that amount at that resolution. This bias declined with each level of map aggregation, being reduced to nearly 15\% for GlobCover and 5\% for MODIS at 100 km. The magnitude of mean absolute error (MAE) was somewhat higher in all cases. The GeoWiki map, in contrast, was the least biased overall, showing just a ~7\% bias at 1 km and near 0 for all other scales of aggregation, although its accuracy (23\% MAE) was only half as good as SA-LC's at 1 km (11\% MAE), which despite its uniform overestimation bias (Fig. 1A) was the most accurate map at aggregation scales $<10$km. Above this, GeoWiki became slightly more accurate, having $<$5\% MAE at 100 km resolution. The reason GeoWiki had relatively poor accuracy at 1 km resolution was due to the high heterogeneous error pattern, which traded between positive and negative residuals over short distances, thereby inflating MAE at this scale.

\subsection{Error and landcover density}
Given that landcover characteristics can influence landcover map quality \cite{see_improved_2015, estes_platform_2015, debats_generalized_2015}, we set out to elucidate the relationship between cropland configuration and map error.  We first calculated the average 1 km reference cropland percentage within each of South Africa's 354 magisterial districts (the finest administrative unit, averaging 3,445 km$^2$; SI), which yielded a landscape-scaled metric of characteristic cropland density, and similarly calculated the district-wise MAE for each test map.

\begin{figure*}[hb]
%\begin{wrapfigure}{c}{1\textwidth}
\centerline{\includegraphics[width=0.95\textwidth]{figures/figure1.pdf}}
\vspace{-0.15 cm}
\caption{(A) Errors in the percent cropland estimates resulting from each of the four test maps relative to the reference map at different scale of pixel aggregation. Rows indicate the test map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas where areas under communal farmlands or permanent tree crops were removed from analysis. (B) The bias (mean error) and accuracy (mean absolute error [MAE]) of each test map at each scale of aggregation, weighted by the percentage cropland in each cell of the reference map. Bias estimates are indicated by the semi-transparent bars, accuracy (lower is more accurate) by the solid bars, with bar colors coded to specific cropland maps.}
\label{afoto1}
\end{figure*}

@@

The magnitude of carbon errors varied as a function of the carbon density of surrounding cover, as demonstrated by the bias statistics (Fig. 3). Bias was near zero when grassland was the adjacent cover type (SI), as its carbon density is nearly the same as cropland. However, when forest was adjacent then bias was a three- to five-fold multiple of cropland map bias (Fig. 1B). At the most extreme, GlobCover's bias was -276\% at 1 km, but even SA-LC and GeoWiki had biases of 22\% and -50\%, respectively. Bias could be substantial even for the least carbon dense vegetation type (sparse), as evidenced by the 15-25\% mean error at 1 km for MODIS and GlobCover under this class.  The mean bias across the different potential adjacent vegetation classes ranged between -20 for GeoWiki and -123\% for GlobCover at 1 km (with MODIS in between these), while SA-LC's average bias was 11\%.  Biases declined fairly rapidly with aggregation, with all datasets having an average (across cover types) bias magnitude of $<$10\% at $\geq$25 km of aggregation, except for GlobCover, which was -12\% at 100 km (SI).  As with cropland percentages, GeoWiki produced the least biased carbon density estimates above 1 km resolution.

In terms of accuracy, MAE values were essentially the same as bias magnitudes, except for GeoWiki's, which were twice as large. The average MAE across vegetation classes was 47\% at 1 km, dropping to $<$10 only with 25 km of aggregation. In contrast, SA-LC's carbon estimates were twice as accurate at 1 km, and were slightly more accurate up to 25 km of aggregation were GeoWiki achieved parity.

@@

We used these same methods  \cite{ramankutty_farming_2008,monfreda_farming_2008} to disaggregate maize harvested area (South Africa's largest crop \cite{estes_projected_2013}) on top of our reference and cropland maps, followed by yields, which were assigned to cells have harvested areas greater than zero. We then used these two layers to calculate maize production, and further aggregated the yield and production grids to 5, 10, 25, 50, and 100 km resolutions before quantifying the bias and accuracy of each test map's yield and production values. In this case, we could not convert cell-wise errors into percentages of the reference map values (because many cells had zero values for one map but not the other), so we calculated bias and accuracy from the map residuals and then normalized their values to the reference map means.

Maize yields disaggregated onto the test maps showed some marked differences relative to the reference map, but only at the margins of the major crop production areas where cropland is sparser (SI). These differences resulted when a yield value was mapped onto a grid cell where the reference map had no harvested area, and thus zero yield. In more densely cropped areas, such discrepancies were less frequent because both the reference and test maps were both likely to have some maize harvested area, and therefore a yield value.  Yield biases were thus fairly low (and accuracy high), with the largest being 20\% for MODIS at 1 km, following by GlobCover with 10\% (Fig. 5). These dropped to $<$10\% with aggregation.

Production biases were generally slightly higher, but still low, for most datasets, with the exception of GlobCover, which had a large underestimation bias of $>$60\% (relative to mean production) at 1 km, which remained above 10\% even at 100 km of aggregation. MODIS production bias was above 20\% at 1 km, but declined to below 10\% at higher levels of aggregation.

In contrast, the accuracy of production estimates was poor. Here all datasets but SA-LC had MAE values of $\geq$30\% below 25 km of aggregation (Fig. 5), reaching as high as 100\% for GlobCover at 1 km, followed by 65\% for MODIS and 45\% for GeoWiki. SA-LC estimated production was most accurate, having between 10-20\% MAE between 1 and 10 km, and $<$10\% at 25 km and higher.  This low accuracy relative to the gridded yield measures relates to the disaggregation process for harvested area, which allocates a fractional value to each pixel, which is itself a fraction. The process of adjusting the gridded values so that their totals match reported statistics does relatively little to correct the map's underlying commission or omission errors, and this constraint in fact appears to shorten the distance between negative and positive residuals (SI), thereby increasing absolute errors.
